A Core Calculus of Metaclasses

### 2004-12-10 metaclasses/paper.tex
Eric's edits to FOOL paper

\documentclass{acmconfbig}

@@

\author{Sam Tobin-Hochstadt \quad Eric Allen \\
\begin{tabular}{c}
\\
Sun Microsystems Laboratories \\
Burlington MA 01803 \\
{\tt \large <first>.<last>@sun.com} \\
\end{tabular}
}

@@

\begin{abstract}
Metaclasses provide a useful method of abstraction for programmers
working with object-oriented languages, but they have not seen the
formal exploration applied to more conventional object-oriented
programming features.  In order to elucidate the structure of
metaclasses and their relationship with static typing, we present a
core calculus for a nominally-typed object-oriented language with
metaclasses and prove type soundness over this core.  This calculus is
presented as an adaptation of Featherweight GJ \cite{FJ}, and is
powerful enough to capture metaclass relationships beyond what are
expressible in common object-oriented languages, including arbitrary
metaclass hierarchies and classes as values.  We also explore the
integration of object-oriented design patterns with metaclasses, and
show how our calculus integrates several such patterns cleanly.
\end{abstract}

\section{Introduction}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships. For example, consider the relationship between
physical quantities, units of measurement, and physical dimensions. It
is obvious that ``length'' is a dimension, and that ``3 feet'' is a
length.  However, modeling this relationship in a conventional
language such as the Java${^{\mbox{\tiny{TM}}}}$ Programming Language
is impossible without attributing to some of these concepts properties
which do not properly belong to them.  It is natural to model {\txt 3
feet} as an instance of class {\txt Length}. But if we were to define
a class {\txt Dimension} and define {\txt Length} to be an instance of
class {\txt Dimension}, then {\txt Length} would not be a class and so
{\txt 3 feet} could not be an instance of {\txt
Length}. Alternatively, we might define {\txt Length} to be a subclass
of {\txt Dimension}. But then we still cannot define {\txt 3 feet} to
be an instance of {\txt Length} because anything that is an instance
of {\txt Length} would also be an instance of class {\txt Dimension}
and {\txt 3 feet} is obviously not a dimension.

The fundamental limitation of conventional object-oriented languages
is that they do not allow a single concept to serve both as a class
and as an instance.  In \cite{FW96}, Welty and Ferrucci lay out a
comprehensive case that this limitation prevents conventional modeling
languages from capturing many important aspects of the world.  In
their primary example, they show the difficulties in modeling a simple
ontology including the notions {\txt Species} and {\txt Eagle}, while
also capturing the relationship between {\txt Eagle} and {\txt Harry},
a particular eagle.  One solution they consider and reject is to
separate the notion of ``eagle'' into two concepts: {\txt the Eagle}
and {\txt Eagle}, with the former an instance of {\txt Species} and
the latter a class whose instances are specific birds.  This approach
is inadequate because it fails to express a relationship between the
notions {\txt the Eagle} and {\txt Eagle}.
%The languages Ferrucci and Welty discuss cannot accommodate multiple
%universes of discourse, in which a single thing may have different meanings in different universes.

In order to overcome these limitations in expressiveness, several
object-oriented languages that are not statically typed, such as
Smalltalk, Python, and Self, allow for more flexible class
relationships \cite{Smalltalk80}, \cite{ArtMOP}, \cite{PythonMeta}. In
the case of Smalltalk and Python, classes are instances of
metaclasses. Static members of a class are modeled as ordinary members
of the corresponding metaclass.  But traditional metaclass systems
have placed important limitations on expressiveness.  For example, in
Smalltalk, the metaclass hierarchy is only two levels deep.  If we
want to represent the relationships ``{\txt A} is an instance of {\txt
B}", ``{\txt B} is an instance of {\txt C}", and ``{\txt C} is an
instance of {\txt D}", we can only do this in Smalltalk if {\txt D} is
the special Smalltalk class {\txt Metaclass}.  This does not allow the
multi-level hierarchies required to model the examples discussed above
\cite{Smalltalk80}.  Self, on the other hand, is a prototype-based
object-oriented language, where there are no classes at all; object
instantiation consists of cloning an existing instance. The members of
the clone can be modified and added to at will, making static checking
difficult.

An extremely expressive system of metaclasses, very similar to ours,
was presented by Cointe in \cite{objvlisp}.  In his paper, he makes a
number of persuasive arguments for the benefits of metaclasses.
In response to complaints that the Smalltalk80 metaclass system was
too complex, he suggested that a more powerful and flexible system, as
he and we present, has the potential to be much simpler.
He claims ``The ObjVlisp model's primary advantage is uniformity.''
We agree, and believe that the treatment of metaclasses presented here
unifies both disparate metaclass systems, and different ad-hoc
mechanisms describing the properties of classes as objects in
languages without metaclasses.

Cointe also quotes from \cite{cardelli:multiple}: ``With respect to
Simula, Smalltalk also abandons ... strong typing, allowing it to
... introduce the notion of meta-classes''.  In this paper, we show
that such a compromise is not necessary, and that static typing can
coexist with the flexibility of metaclasses.

We present and formalize a statically typed calculus with nominal
subtyping for metaclasses where every class can serve as both an
instance of a class and as a class of instances, and there is no
limitation on the levels of nesting of metaclasses. Our work is
motivated by \cite{DimUnits}, which presents a system for integrating
the static checking of dimensions of physical quantities with a
conventional object-oriented type system. In that system, metaclasses
are used extensively to model the type relationships of dimensions.
The notion of metaclass in that system is more general than that of
languages such as Smalltalk, where each class has a unique metaclass
that defines its class members. In this paper, we formalize the core
of that system related to metaclasses and prove it sound.  In the
process, we discover several interesting properties about the
structure that a nominal type system for metaclasses must take.

In the interests of economy and clarity of presentation, in the
following discussion we restrict ourselves to simple examples, and do
not delve in the the complexities of object-oriented analysis, or of
dimension types.  This is certainly not because we believe that
metaclasses are most useful for the development of toy object-oriented
programs.  On the contrary, metaclasses are likely to show their value
in large systems, both in terms of modeling and in the treatment of
design patterns in section \ref{designpatterns}.

The remainder of this paper is organized as follows.  In section
\ref{MCJ} we present a core calculus for metaclasses we dub MCJ.
We introduce the language and describe its key features, as
well as some of the important design choices.  The motivating examples
from the introduction are presented in MCJ in section \ref{examples},
demonstrating the benefits of metaclasses. We discuss in section
\ref{designpatterns} how the need for many object-oriented ``design
patterns'' is obviated through the use of metaclasses. We then
describe its formal properties in section \ref{formal}.  In section
\ref{formalsem} we provide a formal semantics for the language, and in
section \ref{proof} we prove a type soundness result.  In section
\ref{related} we consider related work and in section \ref{future} we
describe conclusions and future directions.

@@

Each class may have both class members and instance members. Class
members define the behavior of the class when used as an
instance. Instance members define the behavior of instances of a
class.  In the abstract syntax presented in this paper, class members
are distinguished from instance members by order, and with the keyword
{\txt class} \footnote{This syntax is not ambiguous since MCJ has no
inner classes.}.  Members of a class definition are laid out as
follows: first, class fields are defined, followed by class methods,
then instance fields and finally instance methods.  For example, the
following class definition includes a class method {\txt m} and an
instance method {\txt n}:

@@

If the method call were instead $\invoke \bb {\mof {}}$, where {\txt
b} denotes an instance of {\txt B}, then the instance methods of first
$\B$ then $\A$ would be examined.

@@

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be
constructed. Non-class instances are naturally constructed with class
(factory) methods \cite{GOF}; in this way a class instance can be
passed as a type parameter and factory methods can be called on the
class instance to construct new instances without stipulating the
class of the constructed instance. This flexibility is similar to that
provided with first-class genericity \cite{NextGen, MixGen}, but
obviates the need for {\txt with} clauses because the bound on the
kind of a type variable stipulates what factory methods can be called
on it.  A class can define class methods with arbitrary signatures
that return new instances of the class, but each such method must
ultimately includes a {\txt new} expression, which can occur only
within the scope of the class whose instance it returns. A {\txt new}
expression looks like a call to method named {\txt new}, that takes a
single argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt new} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt new} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they could be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt new}
expressions.

@@

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  In the world of the
$\lambda$-calculus, such types are known as kinds.  In MCJ, however,
the kind of a class does not capture all of the properties of the
class as a value.  For example, a class may add new class fields or
class methods which are not present in the kind.  Therefore, each
class freely generates a new type, which is the type of the class
considered as a value.  We represent this type with the {\txt typeOf}
operator, which is produces a type that is not also a class.  Since
this type is not a class, it cannot be the superclass or kind of
another class, and it cannot serve as the instantiation of a generic
type parameter.

@@

The name {\txt typeOf} has been used in other contexts to refer to a
runtime operation that determines the dynamic type of a object.
However, despite the name similarity, these functions bear no
relationship to the operator described here.  In MCJ, an application
of {\txt typeOf} is a static type reference.

@@

The standard subclassing relationship, like subtyping, is transitive.
That is, if $\B$ is subclass of $\C$ and $\C$ is a subclass of $\D$,
then $\B$ is a subclass of $\D$.  However, this relationship does not
hold for {\txt instanceof} relationships.  Therefore, if if $\B$ is an
instance of $\C$ and $\C$ is an instance of $\D$, then no judgment
about the relationship of $\B$ to $\D$ can be inferred.

\paragraph{Cyclic class hierarchies}
One property of metaclasses that we have not yet discussed is that
cycles can occur among instance class relationships. The simplest such
cycle is:

@@

Having seen the outlines of MCJ, we return to the motivational
examples discussed earlier.  First, the relationships between
dimensions is easy to capture. \footnote{Of course, dimensions have
many subtleties, not captured here.  All we present is the essence of
the type relationships.} We define

@@

There are at least two substantial problems with this code.  First,
{\txt TheEagle} and {\txt Eagle} have no relationship to each other in
the type system.  One is a singleton class, representing a particular
species, and one is a name for a set of objects, being the members of
that species.  The fundamental nature of species has a two-level
containment relationship, which the type system is failing to express.
This means that our type checker cannot determine what we want, and
cannot help us avoid mistakes.

Another problem, which is in some ways just a symptom of the first
problem, but which bears special attention, is the use of generic
types.  Imagine the following method, where T is a type parameter in
scope:\footnote{Generic methods would be straightforward to add to
MCJ, but we excluded them to simplify the presentation of the type
system, and since they do not add complexity to the proofs.}

@@

Design Patterns, as exemplified in \cite{GOF}, have had a very
substantial impact on the world of object-oriented programming and
beyond.  Design patterns allow programmers to increase the flexibility
and abstraction of their software designs.  However, few design
patterns have been integrated into programming languages.

One of the advantages of our metaclass framework is that it allows
clean expression of several patterns in the language, rather than
adding additional abstraction techniques on top of the language.  For
example, a number of the classic design patterns deal with object
creation, as discussed above in section \ref{privconst}.  It is clear
that the concept of a Factory method is easily (and necessarily)
expressed in MCJ.  However, a number of other patterns also find easy
expression.  The Abstract Factory pattern is a simple application of
inheritance in the metaclass hierarchy.  The Prototype pattern is, as
mentioned in the Design Patterns book, supported in languages with
metaclasses.  Finally, there is no need for the Singleton pattern,
since the class itself can serve as a singleton.  While this is
possible in the Java Programming Language and in \Cpp, it is
undesirable since the static behavior that would need to be used
cannot inherit from another class.  In fact, the inflexibility of
class or static operations is one of the rationales cited for the
Singleton pattern.

@@

Sophisticated macro systems, such as those found in Common Lisp
\cite{steele-90} or Scheme \cite{SyntaxCase} allow for expression of
many forms of abstraction, as do extremely flexible object systems
such as CLOS. However, these solutions add significantly to what a
programmer must understand in order to use the language productively.

An alternative solution explored in this paper is to provide additional
flexibility in the language, which does not allow for arbitrary
additional abstraction, but instead makes the abstractions commonly
needed easy to express.  One does not need the full power of macros to
encode the Factory pattern.  In MCJ, we have presented a language
where several kinds of patterns are easy to express, without adding an
additional language on top of the original.

@@

This is a class named $\C$, with no type arguments, with kind
$\Object$ and superclass $\Object$, which has no fields or methods
whatsoever.  In examples that follow, we omit empty {\txt <>}.  Note
that both the kind and the superclass are $\Object$.  MCJ does not
have a distinguished class $\Class$.  Interestingly, we determined
that a class $\Class$ would need no consideration from the type
system.  Since both $\Class$ and $\Object$ would sit atop the
hierarchy with no contents, we determined there was no need to include
both of them.  In a more substantial language, where $\Object$ might
have methods or fields, there might be a use for a distinguished
$\Class$ class.  However, it would not need any special treatment by
the type system.

@@

Other kinds of expressions are seen in the body of the {\txt Pair}
class.  {\txt new<A,B> (a,b)} is a new expression, which creates an
instance of the enclosing class (here {\txt Pair}).  In the body of
{\txt make} is a reference to the variable {\txt b}, which has the
obvious value.  The final directly expressible expression is the cast,
which works as follows: {\txt (Object)O.make()} is an expression of
type \Object that evaluates to an instance of {\txt O}.

@@

With the above definitions, we are now able to turn to a proof of type
soundness. Given the simplicity we are able to achieve in the
definitions, the proof is not significantly more complicated than the
soundness proof given for Featherweight GJ\ \cite{FJ}.  However, there
are a number of significant lemmas, of which we list the most
important:

@@

The metaclass system present in Python \cite{PythonMeta} is quite
similar to that provided here, with inheritance of instance methods as
class methods from arbitrary other classes as metaclasses.  However,
Python is dynamically-typed, and many of the uses to which Python
metaclasses are put are not possible in a statically typed language.
The work on static type systems for Python has not included
metaclasses \cite{Starkiller}.

Several type systems have also been proposed for languages with
metaclasses, including the Strongtalk language \cite{Strongtalk},
which turns Smalltalk into a structurally-typed language with
extensive static checking.  However, because the Smalltalk metaclass
system is so different from the one in MCJ, many of the interesting
aspects of the type system do not carry over.  Furthermore, the
Strongtalk papers do not provide a formal semantics and analysis of
the system.  A formal analysis of inheritance in Smalltalk is provided
in \cite{cook:thesis} but this again does not consider the hierarchy
of metaclasses presented here.

@@

Cointe \cite{objvlisp} presents a model of metaclasses extremely
similar to that presented here.  In it, he provides several
overlapping motivations to our own.  One is to regularize the
metaclass system of Smalltalk, and another is to enable additional
programming flexibility.  To this we add modeling freedom, and a
relation to static methods in more recent OO languages.  Cointe's
work, however, does not provide a formal model, so it is difficult to
determine the exact relationship between the systems.  Additionally,
his work is in a untyped setting (Lisp and Smalltalk) and thus the
safety theorems proved here are not possible.

@@

We would like to thank Jan-Willem Maessen for his extremely helpful
feedback on the formal rules presented in this paper.  We also thank
David Chase for not believing in us unless we were right, Victor
Luchangco for many valuable discussions, and Guy Steele for his many
helpful comments.

>>>>

\documentclass{acm-sigplan}

@@

\authorinfo{Sam Tobin-Hochstadt}{Northeastern University}{samth@ccs.neu.edu}
\authorinfo{Eric Allen}{Sun Microsystems Laboratories}{eric.allen@sun.com}

%\authorinfo{Sam Tobin-Hochstadt \quad Eric Allen \\
%\begin{tabular}{c}
%\\
%Sun Microsystems Laboratories \\
%Burlington MA 01803 \\
%{\tt \large <first>.<last>@sun.com} \\
%\end{tabular}
%}

@@

\begin{abstract}
Metaclasses provide a useful mechanism for abstraction in
object-oriented languages.  But most languages supporting metaclasses
have imposed severe restrictions on their use. Typically, a metaclass
is allowed to have only a single instance and all metaclasses are
required to share a common superclass \cite{objvlisp}. In addition,
few languages supporting metaclasses have included a static type
system, and none have included a type system with nominal subtyping
(i.e., subtyping as defined in mainstream languages such as the
Java${^{\mbox{\tiny{TM}}}}$ Programming Language or C\#).

In order to elucidate the structure of
metaclasses and their relationship with static types, we present a
core calculus for a nominally-typed object-oriented language with
metaclasses and prove type soundness over this core. 
To our knowledge, no previous formalization of metaclasses in a 
language with nominal subtyping exists.
This calculus is
presented as an adaptation of Featherweight GJ \cite{FJ}, and is
powerful enough to capture metaclass relationships beyond what are
expressible in common object-oriented languages, including arbitrary
metaclass hierarchies and classes as values.  We also explore the
integration of object-oriented design patterns with metaclasses, and
show how our calculus integrates several such patterns cleanly.
\end{abstract}

\section{Introduction}
One of the stated benefits of object-oriented languages is their
ability to model aspects of the world in the object hierarchy.
However, most such languages are unable to model many simple and
common relationships. For example, consider the relationship between
physical quantities, units of measurement, and physical dimensions. It
is obvious that ``length'' is a dimension, and that ``3 feet'' is a
length.  However, modeling this relationship in a conventional
language such as the Java Programming Language is difficult without
attributing to some of these concepts properties which do not properly
belong to them.  It is natural to model {\txt 3 feet} as an instance
of a class {\txt Length}. But if we were to define a class {\txt
Dimension} and define {\txt Length} to be an instance of class {\txt
Dimension}, then {\txt Length} could not be a class and so {\txt 3
feet} could not be an instance of {\txt Length}. Alternatively, we
might define {\txt Length} to be a subclass of {\txt Dimension}. But
then we still cannot define {\txt 3 feet} to be an instance of {\txt
Length} because anything that is an instance of {\txt Length} would
also be an instance of class {\txt Dimension} and {\txt 3 feet} is
obviously not a dimension. The key limitation preventing us from
modeling this domain properly is that we cannot define a single
concept to serve both as a class and as an instance of a class.

Another example is provided by Welty and Ferrucci in \cite{FW96}, where
they show the difficulties in modeling a simple
ontology including the notions {\txt Species} and {\txt Eagle}, while
also capturing the relationship between {\txt Eagle} and {\txt Harry},
a particular eagle.  One solution they consider and reject is to
separate the notion of ``eagle'' into two concepts: {\txt the Eagle}
and {\txt Eagle}, with the former an instance of {\txt Species} and
the latter a class whose instances are specific birds.  This approach
is inadequate because it fails to express a relationship between the
notions {\txt the Eagle} and {\txt Eagle}.
Again, the limitation is that we cannot define a class as an instance of another class.
%Welty and Ferrucci proceed to explain how this limitation prevents conventional modeling
%languages from capturing many important aspects of the world.  
%The languages Ferrucci and Welty discuss cannot accommodate multiple
%universes of discourse, in which a single thing may have different meanings in different universes.

In order to overcome this limitation in expressiveness, several
object-oriented languages that are not statically typed, such as
Smalltalk, Python, and Self, allow for more flexible object
relationships \cite{Smalltalk80}, \cite{ArtMOP}, \cite{PythonMeta}. In
the case of Smalltalk and Python, classes are instances of
metaclasses. Static members of a class are modeled as ordinary members
of the corresponding metaclass.  But in both of these languages there
are important limitations on expressiveness.  For example, in
Smalltalk, the metaclass hierarchy is only two levels deep.  If we
want to represent the relationships ``{\txt A} is an instance of {\txt
B}", ``{\txt B} is an instance of {\txt C}", and ``{\txt C} is an
instance of {\txt D}", we can only do this in Smalltalk if {\txt D} is
the special Smalltalk class {\txt Metaclass}.  This does not allow the
multi-level hierarchies required to model the examples discussed above
\cite{Smalltalk80}.  Self, on the other hand, is a prototype-based
object-oriented language, where there are no classes at all; object
instantiation consists of cloning an existing instance. The members of
the clone can be modified and added to at will, dispensing with class
relationships entirely and making static checking difficult.

A more expressive metaclass system in ObjVlisp, described informally
by Conte in \cite{objvlisp}.  In ObjVlisp, there is no restriction on
the number of instances of a metaclass, and metaclasses are not
required to share a common superclass. The instance methods of a
metaclass are inherited as the static members of its instances.  Conte
makes a number of arguments for the benefits of generalized
metaclasses.  In response to complaints that the Smalltalk80 metaclass
system is too complex, he suggests that a more powerful and flexible
system has the potential to be much simpler.
%, arguing that ``The ObjVlisp model's primary advantage is uniformity.''
However, ObjVlisp does not include a static type system. 
%Cointe quotes from \cite{cardelli:multiple}: ``With respect to
%Simula, Smalltalk also abandons ... strong typing, allowing it to
%... introduce the notion of meta-classes''.   
In this paper, we show that this limitation is not necessary. We
present a semantics for a statically typed calculus of metaclasses
where, as in ObjVlisp, every class can serve as both an instance of a
class and as a class of instances, and there is no limitation on the
levels of nesting of metaclasses. In the process, we also reveal
several interesting properties about the structure that a nominal type
system for metaclasses must take. We then present a formal operational
semantics for this calculus, as well as a proof of type soundness.

Our work is motivated by \cite{DimUnits}, which presents a system for
integrating the static checking of dimensions of physical quantities
in MetaGen, an extension of the Java Programming Language with support
for metaclasses. In MetaGen, metaclasses are used to model the type
relationships of dimensions.  As with ObjVlisp, the notion of a
metaclass in MetaGen is more general than that of languages such as
Smalltalk, but, unlike ObjVlisp, MetaGen includes a static type system
\cite{DimUnits}. The calculus in this paper can be viewed as a
formalization of a purely functional core of MetaGen.

In the interests of economy and clarity of presentation, in the
following discussion we restrict ourselves to simple examples, and do
not delve into the complexities of object-oriented analysis with metaclasses.  
This is certainly not because we believe that
metaclasses are most useful for the development of toy object-oriented
programs.  On the contrary, metaclasses are most likely to show their value
in large systems, both in terms of modeling and in the treatment of
design patterns in section \ref{designpatterns}.

The remainder of this paper is organized as follows.  In section
\ref{MCJ} we present a core calculus for metaclasses we dub MCJ.
We introduce the language and describe its key features, as
well as some of the important design choices.  The motivating examples
from the introduction are presented in MCJ in section \ref{examples},
demonstrating the benefits of metaclasses. We discuss in section
\ref{designpatterns} how the need for many object-oriented ``design
patterns'' is obviated through the use of metaclasses. We then
describe the formal properties of MCJ in section \ref{formal}.  In section
\ref{formalsem} we provide a formal semantics for the language, and in
section \ref{proof} we prove a type soundness result.  In section
\ref{related} we consider related work and in section \ref{future} we
describe conclusions and future directions.

@@

Each class may have both class members and instance members. Class
members define the behavior of the class when used as an
instance. Instance members define the behavior of instances of a
class.  In the abstract syntax presented in this paper, class members
are distinguished from instance members by order, and with the keyword
{\txt class} \footnote{This syntax is not ambiguous because MCJ has no
inner classes.}.  Members of a class definition are laid out as
follows: first, class fields are defined, followed by class methods,
then instance fields and finally instance methods.  For example, the
following class definition includes a class method {\txt m} and an
instance method {\txt n}:

@@

If the method call were instead $\invoke \bb {\mof {}}$, where {\txt
b} denotes an ordinary (non-class) instance of {\txt B}, then the
instance methods of first $\B$ then $\A$ would be examined.

@@

One benefit of a metaclass system is that constructors no longer need
to play such a central role in the language. Class instances exist at
the beginning of a program execution; they need not be
constructed. Non-class instances are naturally constructed with class
(factory) methods \cite{GOF}; in this way a class instance can be
passed as a type parameter and factory methods can be called on the
class instance to construct new instances without stipulating the
class of the constructed instance. This flexibility is similar to that
provided with first-class genericity \cite{NextGen, MixGen}, but
obviates the need for {\txt with} clauses because the bound on the
kind of a type variable stipulates what factory methods can be called
on it.  A class can define class methods with arbitrary signatures
that return new instances of the class, but each such method must
ultimately include a {\txt new} expression, which can occur only
within the scope of the class whose instance it returns. A {\txt new}
expression looks like a call to method named {\txt new}, that takes a
single argument for each type parameter and each instance field of the
syntactically enclosing class. The result of a {\txt new} expression
is a new instance of the enclosing class whose type parameters are
instantiated with, and whose fields are initialized with, the given
arguments. In our formal semantics, {\txt new} expressions are
annotated with the name of the enclosing class. These annotations need
not be added by a programmer; they could be added easily by a
straightforward syntactic preprocessing over the program, since they
can be determined merely by the lexical scopes of the {\txt new}
expressions.

@@

Given that class references can be used as expressions, it is natural
to ask: what is the type of a class reference?  
%In the world of the
%$\lambda$-calculus, such types are known as kinds.  
In MCJ, however, the kind of a class does not capture all of the
properties of the class as a value.  For example, a class may add new
class fields or class methods which are not present in the kind.
Therefore, each class freely generates a new type, which is the type
of the class considered as a value.  We represent this type with the
{\txt typeOf} operator, which takes a class instantiation and produces
a type that is not also a class.  Because this type is not a class, it
cannot be the superclass or kind of another class, and it cannot serve
as the instantiation of a generic type parameter.

@@

\hspace{-.43cm}where we use the convention that {\txt A:B} denotes
that {\txt A} has type {\txt B}, and {\txt A<:B} denotes that {\txt A}
is a subtype of {\txt B}.

The name {\txt typeOf} has been used in other contexts to refer to a
runtime operation that determines the dynamic type of a object.
However, despite the name similarity, this function bears no
relationship to the operator described here.  In MCJ, an application
of {\txt typeOf} is a static type reference.

@@

The standard subclassing relationship, like subtyping, is transitive.
That is, if $\B$ is a subclass of $\C$ and $\C$ is a subclass of $\D$,
then $\B$ is a subclass of $\D$.  However, this relationship does not
hold for {\txt instanceof} relationships.  In general, if $\B$ is an
instance of $\C$ and $\C$ is an instance of $\D$, no judgment
about the relationship of $\B$ to $\D$ can be inferred.

Another difference between subclassing and {\txt instanceof}
relationships is that cycles can occur among instance class
relationships. The simplest such cycle is:

@@

Having seen this outline of MCJ, we return to the motivational
examples discussed earlier.  First, the relationships between
dimensions is easy to capture. \footnote{Of course, dimensions have
many subtleties, not captured here.  All we present is the essence of
the type relationships.} We define

@@

There are at least two substantial problems with this code.  First,
{\txt TheEagle} and {\txt Eagle} have no relationship to each other in
the type system.  One is a singleton class, representing a particular
species, and one is a name for a set of objects, being the members of
that species.  The fundamental nature of species has a two-level
containment relationship, which the type system fails to express.
This means that our type checker cannot determine what we want and
cannot help us avoid mistakes.

Another problem, which is in some ways just a symptom of the first
problem but which bears special attention, is the use of generic
types.  Imagine the following method, where T is a type parameter in
scope:\footnote{Generic methods would be straightforward to add to
MCJ, but we exclude them to simplify the presentation of the type
system and to focus on the important interactions of metaclasses with
other language features.}

@@

Design Patterns, as exemplified in \cite{GOF}, have had a
substantial impact on the world of object-oriented programming and
beyond.  Design patterns allow programmers to increase the flexibility
and abstraction of their software designs.  However, few design
patterns have been integrated into programming languages.

One of the advantages of our metaclass framework is that it allows
clean expression of several common design patterns in the language,
rather than requiring that they be expressed with abstraction
techniques on top of the language.  For example, a number of the
classic design patterns deal with object creation, as discussed above
in section \ref{privconst}.  It is clear that the concept of a Factory
method is easily (and necessarily) expressed in MCJ.  However, a
number of other patterns also find easy expression.  The Abstract
Factory pattern is a simple application of inheritance in the
metaclass hierarchy.  The Prototype pattern is, as mentioned in the
Design Patterns book, supported in languages with metaclasses.
Finally, there is no need for the Singleton pattern; a class itself
can serve as a singleton.  While classes can be used as limited
singletons in the Java Programming Language and in \Cpp, doing so is
undesirable because the static behavior of a class cannot inherit from
another class.  In fact, the inflexibility of class or static
operations is one of the rationales cited for the Singleton pattern.

@@

Sophisticated macro systems, such as those found in Common Lisp
\cite{steele-90} or Scheme \cite{SyntaxCase} allow for expression of
many forms of abstraction, as do extremely flexible object systems
such as CLOS. However, these solutions add significantly to what a
programmer must understand in order to use the language productively.
In contrast, the inclusion of metaclasses does not allow for arbitrary
additional abstraction, but instead makes the abstractions commonly
needed easy to express.  One does not need the full power of macros to
encode the Factory pattern.  In MCJ, we have presented a language
where several kinds of patterns are easy to express, without adding an
additional language on top of the original.

@@

This is a class named $\C$, with no type arguments, with kind
$\Object$ and superclass $\Object$, which has no fields or methods
whatsoever.  In examples that follow, we omit empty {\txt <>}.  Note
that both the kind and the superclass are $\Object$.  MCJ does not
have a distinguished class $\Class$.  Interestingly, we determined
that a class $\Class$ would need no consideration from the type
system.  Because both $\Class$ and $\Object$ would sit atop the
hierarchy with no contents, there is no need to include
both of them.  In a more substantial language, where $\Object$ might
have methods or fields, there might be a use for a distinguished
$\Class$ class, but it would not need any special treatment by
the type system.

@@

Other kinds of expressions are seen in the body of the {\txt Pair}
class.  {\txt new<A,B> (a,b)} is a new expression, which creates an
instance of the enclosing class (here {\txt Pair}).  In the body of
{\txt make} is a reference to the variable {\txt b}, which has the
obvious value.  The final form of expression writable by the
programmer is the cast, with the usual semantics. For example, {\txt
(Object)O.make()}\ is an expression of type \Object that evaluates to
an instance of {\txt O}.

@@

With the above definitions, we are now able to turn to a proof of type
soundness. Given the simplicity we are able to achieve in the
definitions, the proof is not significantly more complex than the
soundness proof given for Featherweight GJ\ \cite{FJ}.  However, there
are a number of significant lemmas, of which we list the most
important. The full proof is available at:

\vspace{.1cm}
\verb|http://research.sun.com/projects/plrg/|

@@

The metaclass system present in Python \cite{PythonMeta} is
similar to that provided here, where classes inheritance the instance methods
of their metaclasses as class methods.  However,
Python is dynamically-typed, and many of the uses to which Python
metaclasses are put are not possible in a statically typed language.
The work on static type systems for Python has not included
metaclasses \cite{Starkiller}.

Several type systems have also been proposed for languages with
metaclasses, including the Strongtalk language \cite{Strongtalk},
which turns Smalltalk into a structurally-typed language with
static checking.  However, because the Smalltalk metaclass
system is so different from the one in MCJ, many of the interesting
aspects of the type system do not carry over.  Furthermore, the
Strongtalk papers do not provide a formal semantics and analysis of
the system.  A formal analysis of inheritance in Smalltalk is provided
in \cite{cook:thesis} but this again does not consider the hierarchy
of metaclasses presented here.

@@

Cointe \cite{objvlisp} presents a model of metaclasses
similar to that presented here, but for a dynamically typed object-oriented
language based on Lisp.  In it, he provides several
overlapping motivations to our own.  One is to regularize the
metaclass system of Smalltalk, and another is to enable additional
programming flexibility.  To this we add modeling freedom, and a
relation to static methods in more recent OO languages.  Cointe's
work, however, does not provide a formal model, so it is difficult to
determine the exact relationship between the two systems.  Additionally,
his work is in a untyped setting (Lisp and Smalltalk) and thus the
safety theorems proved here are not possible.

@@

We would like to thank Jan-Willem Maessen for his 
feedback on the formal rules presented in this paper.  We also thank
David Chase for not believing in us unless we were right, Victor
Luchangco for many valuable discussions, and Guy Steele for his many
helpful comments.
